---
title: "Leuk_18_full_results"
author: "Piotr Stomma"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What this file contains

In this file, summary tables of all methods tested over all thresholding strategies are shown.
In particular, for each combination of tested number of genes used, thresholding strategy and algorithm, we show its score in terms of Q1 and Q3 of Adjusted Rand Index (ARI) over 100 trials of UMAP embeddings on the `Leuk_18` dataset.

Based on that, we also calculate near the end the thresholding strategy for each combination of algorithm and gene number used that was unambiguously better from the other ones according to the IQR of ARI.
If one wishes to run this code then one should do it after completing `s2b_cl_summary.R` script.


```{r calc, message=FALSE, warning=FALSE}
all_stuff<-readRDS("../leuk_clusters_results_aggregated.rds")
flatten_algo_summary<-function(algo_mat,thr_variant){
  
        algo<-colnames(algo_mat)
        colnames(algo_mat)<-NULL
        data.frame(algorithm= rep(algo,nrow(algo_mat)),
          score=as.vector(t(algo_mat)),
          thr=thr_variant)
}


do.call(rbind,lapply(seq_along(all_stuff), function(n_vc)
    { all_stuff[[n_vc]]-> vc_variant
        rbind(
        flatten_algo_summary(vc_variant$ARI$none, "none"),
        flatten_algo_summary(vc_variant$ARI$complexity, "complexity"),
        flatten_algo_summary(vc_variant$ARI$mst, "mst")
        )-> vc_df
        vc_df$v_cutoff<- n_vc
        vc_df
          }
  ))-> ARI_df

library(dplyr)

v_cutoff_map<- c(
  100,500,1000,
                      2000,3500, 5000,
                    7500, 10000, 12500,
                    15000, 17500, 20000
)

ARI_df$v_cutoff<- v_cutoff_map[ ARI_df$v_cutoff ]

cutoffs_left<- c(500, 1000,2000,
                 5000,10000,20000)

ARI_df<- ARI_df[ ARI_df$v_cutoff %in% cutoffs_left, ]
simil_based<- c("clq_modularity","clq_MSTDunn_sT",      
                "infomap","label_prop",       
                "leiden","affinity_prop",    
                "hcl_single","hcl_complete",     
                "hcl_average")


metric_based<- c("hcl_ward.D2",
                 "k_means","db_scan")

tab_unthr<- setdiff(c(metric_based, simil_based), c("clq_modularity","clq_MSTDunn_sT"))
                    
summarise_ARI<- function(DF, subgroup, q=0.75) {
  DF[subgroup,] %>% group_by(v_cutoff,algorithm) %>%
  summarise(Q1=quantile(score,0.25),
            Qu=quantile(score,q)) %>% as.data.frame() -> df_long
  colnames(df_long)[[4]]<- paste0("Q", ifelse(q==0.75,3,q))
  reshape(df_long, direction="wide", timevar="algorithm",
          idvar="v_cutoff")
}

simil_based_pure= setdiff(simil_based, c("clq_modularity","clq_MSTDunn_sT"))
clq_based<-  c("clq_modularity","clq_MSTDunn_sT")
mask_Sall<- (ARI_df$algorithm %in% simil_based_pure)
mask_clq<- (ARI_df$algorithm %in% clq_based) & 
            (ARI_df$thr!="none") 

simil_cmpl_mask<- (ARI_df$algorithm %in% simil_based) &
                (ARI_df$thr=="complexity")
simil_mst_mask<- (ARI_df$algorithm %in% simil_based) &
                (ARI_df$thr=="mst")
unthr_mask<- (ARI_df$algorithm %in% tab_unthr) &
              (ARI_df$thr=="none")
summarise_ARI(ARI_df, simil_cmpl_mask)-> cmpl_table
summarise_ARI(ARI_df, simil_mst_mask)-> mst_table
summarise_ARI(ARI_df, unthr_mask)-> unthr_table



display_ARI_table<- function(colwise_tab){
 rowtab<- t(round(colwise_tab,2))
 colnames(rowtab)<- rowtab[1,]
 rowtab<- rowtab[-1,]
 as.data.frame(rowtab)
}
```
## Complexity thresholding

```{r}
display_ARI_table(cmpl_table)
```

## MST thresholding

```{r}
display_ARI_table(mst_table)
```

## unthresholded algorithms & metric based methods

```{r}
display_ARI_table(unthr_table)
```
### Which thresholding is best for similarity based methods?

In the main manuscript, we underlined each table entry if that particular thresholding strategy was best for each combination of gene number used and algortihm.
This table repeats this information, but the code below shows how exactly this criterion was calculated.

```{r, warn=FALSE, message=FALSE}



ARI_df[mask_Sall,] %>% group_by(v_cutoff,algorithm,thr) %>%
  summarise(Q1=quantile(score,0.25),
            Q2=median(score),
            Q3=quantile(score,0.75)) %>% as.data.frame() -> iqr_df
iqr_df[,1:2] %>% unique() -> v_alg
v_alg_thr<-cbind(v_alg,best_thr=rep(NA,nrow(v_alg)),Q2=rep(NA,nrow(v_alg)),IQR=rep(NA,nrow(v_alg)))
for (i in 1:nrow(v_alg)){
  vat<- v_alg[i,]
 (iqr_df$v_cutoff==vat[[1]])&
 (iqr_df$algorithm==vat[[2]])-> sub_mask
  iqr_df[sub_mask,]->chunk
  which((chunk$Q1 > chunk$Q3[c(2,1,2)]) & (chunk$Q1 > chunk$Q3[c(3,3,1)]) )-> x
  if (length(x)) {
    (v_alg_thr$v_cutoff==chunk[x,]$v_cutoff) &
    (v_alg_thr$algorithm==chunk[x,]$algorithm) -> v_alg_mask
    stopifnot(sum(v_alg_mask)==1)
    v_alg_thr$best_thr[[which(v_alg_mask)]]= chunk[x,]$thr
    v_alg_thr$Q2[[which(v_alg_mask)]]= chunk[x,]$Q2
    v_alg_thr$IQR[[which(v_alg_mask)]]= chunk[x,]$Q3 - chunk[x,]$Q1
    
  }
}
print("unambigiously best thr per tested variant:")
v_alg_thr
#ARI_df[mask_clq,]
ARI_df[mask_clq,] %>% group_by(v_cutoff,algorithm,thr) %>%
summarise(Q1=quantile(score,0.25),
          Q2=median(score),
          Q3=quantile(score,0.75)) %>% as.data.frame() -> iqr_df_2
iqr_df_2[,1:2] %>% unique() -> v_alg_2
v_alg_thr_2<-cbind(v_alg_2,best_thr=rep(NA,nrow(v_alg_2)),Q2=rep(NA,nrow(v_alg_2)),
                   IQR=rep(NA,nrow(v_alg_2)))
for (i in 1:nrow(v_alg_2)){
  vat<- v_alg_2[i,]
  (iqr_df_2$v_cutoff==vat[[1]])&
    (iqr_df_2$algorithm==vat[[2]])-> sub_mask
  iqr_df_2[sub_mask,]->chunk
  which((chunk$Q1 > chunk$Q3[c(2,1)]) )-> x
  if (length(x)) {
    (v_alg_thr_2$v_cutoff==chunk[x,]$v_cutoff) &
      (v_alg_thr_2$algorithm==chunk[x,]$algorithm) -> v_alg_mask
    stopifnot(sum(v_alg_mask)==1)
    v_alg_thr_2$best_thr[[which(v_alg_mask)]]= chunk[x,]$thr
    v_alg_thr_2$Q2[[which(v_alg_mask)]]= chunk[x,]$Q2
    v_alg_thr_2$IQR[[which(v_alg_mask)]]= chunk[x,]$Q3 - chunk[x,]$Q1
    
  }
}
print("unambigiously best thr per tested variant of CliqueSimNet:")
v_alg_thr_2


```

